Lynott et al. (2014) - Potential error in statistics reporting

### 2017-11-17 manuscript.Rmd
new version manuscript

---
title             : "Equivalence Testing for Psychological Science"
shorttitle        : "Equivalence Testing"

@@

abstract: >
  Psychologists must be able to test both for the presence of an effect and for the the absence of an effect. In addition to testing against zero, researchers can use the Two-One-Sided Tests (TOST) procedure to test for *equivalence* and reject the presence of a meaningful effect. TOST can be used to determine if  an observed effect is surprisingly small, given that a true effect at least as large as the SESOI exists. We explain a range of approaches to determine the SESOI in psychological science, and provide detailed examples of how equivalence tests should be performed and reported. Equivalence tests enable researchers to falsify predictions about the presence, and declare the absence, of meaningful effects.

keywords          : "Equivalence Testing, NHST, power, TOST"
wordcount         : "X"

@@

Alternatively, the researcher could decide that even if the true ratio in the population is not exactly 0.015, the null hypothesis should consist of a range of values around the difference in proportions of 0.015 that can be considered trivially small. The researcher could for example test if the difference is smaller than -0.005, or larger than 0.035. This test against two bounds, with $H_0$ being a range rather than one value (see Figure 1B), is known as a *minimal effects test* [@Murphy2014].

*Equivalence tests* can be seen as the opposite of minimal effects tests: They examine whether the presence of effects that are large enough to matter can be *rejected* (see Figure 1C). In our example, the researcher can perform an equivalence test to test whether the gender difference in participants is *not* large enough to be considered meaningful: the *smallest effect size of interest* (SESOI). After an extensive discussion with experts, the researcher decides that as long as the difference in proportions is not more extreme than 6%, the gender difference is too small to care about. Given an expected true difference in the population of 0.015, the researcher will test if the observed difference falls outside the boundary values of -0.055 and 0.075. If differences more extreme than these boundary values can be rejected in two one-sided tests (TOST), the researcher will conclude statistical equivalence, the gender difference will be considered trivially small, and no money will be spent on addressing a gender difference in participation.

In any one-sided test, for an alpha level of 5%, we can reject $H_0$ when the 90% confidence interval around the observed estimate is in the predicted direction, and does not contain the value tested against (e.g., 0). In TOST, the first one-sided test is used to test against values smaller than the lower bound ($\Delta_{L}$), and the second one-sided test is used to test against values larger than the upper bound ($\Delta_{U}$). It is not necessary to control for multiple tests when using the TOST procedure because both tests need to be statistically significant to conclude equivalence. This means that when reporting a TOST result, authors can suffice by reporting the one-sided test with the smaller test parameter (e.g. $t$ or $r$) and thus the larger $p$-value. Statistical equivalence can be concluded when the largest of the two $p$-values is smaller than alpha.

Note that in this paper we will control the Type 1 error rate at 5% for all examples, mirroring the studies we reanalyze (but we recommend to set and justify the Type 1 error rate in original research based on the goals researchers have [@Lakens2017b]. It might be easiest to think of an equivalence test as checking whether the confidence interval excludes the SESOI, but for any given study an equivalence test could also be conceptualized as determining whether effect sizes or test statistics are closer to zero than some critical value, or even whether the *p*-value for a NHST is larger than some *p*-value bound.

@@

In this article we provide several reproducible examples of equivalence tests which illustrate the procedure in easy-to-use free software (R, jamovi, and a spreadsheet). But first, we discuss different approaches to determining the SESOI for psychological research, which determines the statistical question an equivalence test answers.

@@

TOST is performed against equivalence bounds that are considered the smallest effect size of interest. The SESOI can sometimes be determined objectively, for example based on just noticeable differences. In lieu of objective justifications, the SESOI should ideally be based on a cost-benefit analysis. Since both costs and benefits are necessarily subjective, the SESOI will vary across researchers, fields, and time. The goal of setting a SESOI is to clearly justify why designing a study that has a high probability of rejecting effects more extreme than the specified equivalence bounds contributes to our knowledge base. A SESOI should be chosen such that inferences based on it answer a meaningful question. Although we use bounds that are symmetric around 0 in all examples in this manuscript for equivalence tests (e.g., $\Delta_{L} = -0.3$, $\Delta_{U} = 0.3$), it is also possible to use asymmetric bounds, for example based on an explicit cost-benefit analysis (e.g., $\Delta_{L} = -0.2$, $\Delta_{U} = 0.3$).

@@

Second, researchers can determine the SESOI based on previous work in the literature. Ideally, researchers would always specify their SESOI in studies they report, but this is not yet common practice. It is thus up to researchers who build on earlier work to decide which effect size is too small to be meaningful, given an earlier study. @Simonsohn2015 recently proposed to set the SESOI to the effect size that an earlier study on the same effect would have had 33% power to detect. With this so-called 'small telescopes' approach, the equivalence bounds are thus determined primarily based on the sample size in the original study. For example, consider a study in which `r N` participants answered a question, and the results were analyzed with a one-sample t-test. For a two-sided test with an alpha of `r alpha`, this test had 33% power to detect an effect of $d = `r printnum(pwr.t.test(n=N,sig.level=alpha,power=0.33,type="one.sample",alternative="two.sided")$d)`$. Another example of how previous research can be used to determine the SESOI can be found in @Kordsmeyer2017, who defined the SESOI based on the mean of effect sizes reported in the literature. Thus, they examined whether their replication study could reject effects as large or larger than had on average been reported in the literature. Given random variation and bias in the literature, a more conservative approach could be to use the lower end of a confidence interval around the meta-analytic effect size estimate [cf. @Perugini2014].

Another justifiable choice when basing the SESOI on previous work is to use the smallest observed effect size that could have been statistically significant in a previous study. In other words, we decide that effects that could not have yielded a $p < \alpha$ in an original study will not be considered meaningful in the replication study either, even if they were found to be statistically significant. Based only on the alpha level and the sample size, we can calculate the critical test value (e.g., *t*, *F*, *Z*). This critical test value can be transformed to a standardized effect size (e.g., $d = t \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } }$), which can thus be interpreted as the *critical effect size*.^[This will typically, although not always, correspond to the effect size the study had 50% power to detect [see @Lenth2007]. This procedure will result in equivalence bounds that are substantially wider than the ones obtained using the small telescopes approach, which gives the effect size a study had 33% power to detect.] All observed effect sizes smaller than the critical effect size would not have been statistically significant in the original study, given the alpha and sample size of that study. By setting the SESOI to the critical effect size, an equivalence test can reject all observed effect sizes that could have been detected in an earlier study.

@@

Importantly, an equivalence test based on this approach does not answer any theoretical question (after all, the equivalence bounds are not based on any theoretical predictions) but it answers a *resource* question: Can effects be rejected that could reliably be detected given a specific sample size? In this case, concluding equivalence would suggest that effects which could reasonably be detected with a sample size of 100 can be rejected, and that future studies probably need larger sample sizes to examine a specific question. Whether or not the answer such an equivalence test provides is interesting depends on the sample sizes used in the study, and in part on how typical the chosen sample size is for the research area. An equivalence test based on 15 observations rejecting a SESOI of $d = 1$ cannot be expected to substantially increase our knowledge, given that most effects in psychology are substantially smaller than these bounds and require larger sample sizes to be detected. By transparently reporting the effects that can be detected and rejected based on the study design, researchers can communicate the information their study contributes, and provide a starting point for a discussion about what the size of a reasonable SESOI may be.

@@

The SESOI, and thus the equivalence bounds, can be set in terms of standardized effect sizes (e.g. a Cohen’s $d$ of 0.5) or as a raw mean difference (e.g. 0.5 points on a 7-point scale). The key difference is that equivalence bounds set in raw differences are independent of the standard deviation, while equivalence bounds set as standardized effects are dependent on the standard deviation (since they are calculated as $\frac {X _ { 1} - X _ { 2} } {S D}$. The observed standard deviation randomly varies across samples. In practice, this means that when using standardized differences as bounds (e.g., $d = 0.35$), the equivalence test depends on the standard deviation in the sample. The equivalence bound for a raw mean difference of 0.5 equals a standardized equivalence bound of $d = 0.5$ when the standard deviation is 1, but a standardized equivalence bound of $d = 1$ when the standard deviation is 0.5.

@@

@Banerjee2012 reported that `r orig.N` participants who had been asked to describe an unethical deed from their past judged the room to be darker than participants who had been asked to describe an ethical deed ($t(`r orig.df`)= `r orig.t`$, $p= `r printp(orig.p)`$, $d= `r printnum(orig.d)`$). A close replication by @Brandt2014 with `r rep.n1+rep.n2` participants found no significant effect ($M_{unethical}= `r rep.m1`$, ${SD}_{unethical}= `r rep.sd1`$, $M_{ethical}=`r rep.m2`$, ${SD}_{ethical}=`r rep.sd2`$, $t(`r printnum(rep.df)`)=`r printnum(rep.t)`$, $p=`r printp(rep.p)`$, $d=`r printnum(rep.d)`$). Using the small telescopes approach, we can calculate the effect size the original study had 33% power to detect --- $d = `r printnum(d.33)`$ --- and use this as the SESOI. A TOST procedure for Welch's *t*-test for independent samples and equivalence bounds of ${\Delta}_L=`r printnum(-d.33)`$ and ${\Delta}_U=`r printnum(d.33)`$ reveals that the effect observed in the replication study is statistically equivalent, $t(`r printnum(Brandt$TOST_df)`)=`r printnum(min(Brandt$TOST_t1, Brandt$TOST_t2))`$, $p=`r printp(max(Brandt$TOST_p1, Brandt$TOST_p2))`$ (Figure 2B).

@@

@Hyde2008 report that effect sizes for gender differences in mathematics tests across the 7 million students in the US represent trivial differences, which the authors specify as effect sizes smaller than $d = `r printnum(sesoi)`$. For example, in grade 3 the difference is $d = 0.04$, with a standard error of 0.002. When we perform equivalence tests on the meta-analytic effect sizes of IQ difference for grades 2 to 11 (using an alpha level of `r printp(alpha)` to correct for multiple comparisons), using equivalence bounds of $\Delta_{L} = -0.1$ and $\Delta_{U} = 0.1$, we see that indeed, all effect size estimates are measured with such high precision that they are statistically equivalent and can be considered trivially small, according to the authors’ definition (Figure 2C). In the case of grade 3, $z = `r printnum(max(gr3$TOST_Z1, gr3$TOST_Z2))`$, $p = `r printnum(max(gr3$TOST_p1, gr3$TOST_p2))`$. However, note that all of the effects are also statistically different from zero, as one might expect when there is no random assignment to conditions and samples sizes are very large. For grade 3, $z = `r printnum(gr3.z)`$, $p = `r printnum(gr3.p)`$. This is one way in which the use of equivalence tests improves hypothesis testing procedures in science, by allowing researchers to distinguish between statistical significance and practical significance.

@@

Since this is a replication, we could decide that we are interested in whether the observed effect size is smaller than the smallest effect size that the original study could have detected. This means that we use the critical z value (~1.96 in a two-tailed test with an alpha of 0.05) as equivalence bounds ($\Delta = \pm `r printnum(qnorm(1-0.05/2))`$). To figure out what difference corresponds to a critical z in the original study, we multiply critical z with the standard error.

Since the original study had a clear directional hypothesis,  in this replication study we are only interested in whether receiving a warm pack *increases* the proportion of people who choose a gift for a friend.  This means we can test for inferiority (see Figure 1D) and conclude the absence of a meaningful effect if the observed effect size is reliably *lower* than the SESOI.
We find that we can reject effects larger than  $\Delta_{U}$, $z = `r printnum(Lynott$TOST_z2)`$, $p `r printp(Lynott$TOST_p2)`$ (see Figure 2D). Thus, we can conclude that the statistically non-significant effect is also statistically smaller than our SESOI.

@@

Kahane and colleagues [-@Kahane2015] investigate how responses to moral dilemmas in which participants have to decide whether or not they would sacrifice one person's life to save several other lives relate to other indicators of moral orientation. Traditionally, greater endorsement for sacrificing one life to save several others has been interpreted as a more "utilitarian" moral orientation (i.e. a stronger concern for the greater good). Kahane et al. contest this interpretation in a number of studies. In Study 4, they compare the traditionally used dilemmas with a new set of new ones in which the sacrifice for the greater good is not another person’s life, but something the participant would have a partial interest in(e.g. buying a new mobile phone vs. donating the money to save lives in a distant country) and find no significant correlation between the two ($r(`r Kahane.N-2`) = `r printnum(Kahane.r.GG)`$, $p = `r printp(Kahane.p)`$), $N = `r Kahane.N`$^[Study 4 in @Kahane2015 had a final sample size of $N = `r nrow(Kahane.data)`$, but due to missing data in `r sum(is.na(Kahane.data))` case, the correlation reported here is based on a sample of $N = `r Kahane.N`$.]). The authors conclude that “a robust association between ‘utilitarian’ judgment and genuine concern for the greater good seems extremely unlikely” (p. 206), given the statistical power their study had to detect meaningful effect sizes.

This interpretation can be formalized by performing an equivalence test for correlations, where equivalence bounds are set to an effect size the study had reasonable power to detect. With `r Kahane.N` participants, the study had 80% power to detect effects of $r = `r printnum(Kahane.r.80)`$. Performing the TOST procedure given bounds of $\Delta_{L} = `r printnum(-Kahane.r.80)`$ and $\Delta_{U} = `r printnum(Kahane.r.80)`$, the result is indeed statistically equivalent, $r(`r Kahane.N-2`) = `r printnum(Kahane.r.GG)`$, $p = `r printp(max(Kahane$TOST_p1, Kahane$TOST_p2))`$. This means that values smaller than $r = `r printnum(-Kahane.r.80)`$ and larger than $r = `r printnum(Kahane.r.80)`$ can be rejected at an alpha level of 5%. If other researchers are interested in examining possible smaller effects, they can design studies with a larger sample size that can detect, or reject, smaller effect sizes.

@@

Equivalence testing is a relatively simple and flexible technique to falsify hypotheses. As long as we can calculate a confidence interval around a parameter estimate, we can compare it to a smallest effect size of interest. The result of an equivalence test can be obtained by mere visual inspection of the CI [@Seaman1998; @Tryon2001], or by performing two one-sided tests (the TOST procedure).

As with any statistical test, the usefulness of the result of an equivalence test depends on the question we ask. The question manifests itself in the bounds we set: Is our data surprising if the true effect lies outside of the bounds? If we test against very wide bounds, observing statistical equivalence can hardly be considered surprising, given that most effects in psychology are small to medium [@Hemphill2003]. Examining the papers citing @Lakens2017a, we found that some researchers state a SESOI without providing a justification for it [@Brown2017; @Schumann2017].^[In these two cases, one could perhaps guess that benchmark values for a Cohen’s $d$ of “medium” size was chosen.] An equivalence test using a SESOI of $d = 0.5$ might very well answer a question the researchers are interested in [for one possible justification based on minimally important differences, see @Norman2003], but researchers should always provide a detailed justification of their chosen SESOI, given that this is in essence a justification for why they ask a specific question from their data. Equivalence bounds should be specified before results are known, ideally as part of a pre-registration [cf. @Piaggio2006]. Researchers might be tempted to set equivalence bounds *after* looking at their data, but like any other results-dependent decision, this "boundary hacking" will increase error rates.

We recommend that researchers by default perform both a null hypothesis significance test and an equivalence test on their data. The biggest challenge for researchers will be specifying the smallest effect size of interest. Psychological theories are usually too vague to derive precise predictions, and if there are no theoretical reference points, natural constraints, or prior studies a researcher can use to define the SESOI for a hypothesis test, any choice will be somewhat arbitrary. In some research lines researchers might simply use equivalence tests to reject consecutively smaller effect sizes by performing studies with increasingly larger sample sizes while controlling error rates, until no one is willing to invest the time and resources needed to examine the presence of even smaller effects. Nevertheless, it is important to realize that not specifying a SESOI for our research questions at all will severely hinder theoretical progress [@Morey2016]. While our initial choices of a smallest effect size of interest may be arbitrary, incorporating equivalence tests in our statistical toolbox will in time contribute to better --- and more falsifiable --- theories.

>>>>

---
title             : "Equivalence Testing for Psychological Research: A Tutorial"
shorttitle        : "Equivalence Testing Tutorial"

@@

abstract: >
  Psychologists must be able to test both for the presence of an effect and for the the absence of an effect. In addition to testing against zero, researchers can use the Two-One-Sided Tests (TOST) procedure to test for *equivalence* and reject the presence of a smallest effect size of interest (SESOI). TOST can be used to determine if an observed effect is surprisingly small, given that a true effect at least as large as the SESOI exists. We explain a range of approaches to determine the SESOI in psychological science, and provide detailed examples of how equivalence tests should be performed and reported. Equivalence tests are an important extension of statistical tools psychologist currently use, and enable researchers to falsify predictions about the presence, and declare the absence, of meaningful effects.

keywords          : "Equivalence Testing, NHST, power, TOST, falsification."
wordcount         : "X"

@@

Alternatively, the researcher could decide that even if the true ratio in the population is not exactly 0.015, the null hypothesis should consist of a range of values around the difference in proportions of 0.015 that can be considered trivially small. The researcher could for example test if the difference is smaller than -0.005, and larger than 0.035. This test against two bounds, with $H_0$ being a range rather than one value (see Figure 1B), is known as a *minimal effects test* [@Murphy2014].

*Equivalence tests* can be seen as the opposite of minimal effects tests: They examine whether the presence of effects that are large enough to be considered meaningful can be *rejected* (see Figure 1C). In our example, the researcher can perform an equivalence test to examine whether the gender difference in participants is *not* as large or larger than the *smallest effect size of interest* (SESOI). After an extensive discussion with experts, the researcher decides that as long as the difference in proportions is not more extreme than 6%, the gender difference is too small to care about. Given an expected true difference in the population of 0.015, the researcher will test if the observed difference falls outside the boundary values (or *equivalence bounds*) of -0.055 and 0.075. If differences more extreme than these boundary values can be rejected in two one-sided tests (TOST), the researcher will conclude statistical equivalence, the gender difference will be considered trivially small, and no money will be spent on addressing a gender difference in participation.

In any one-sided test, for an alpha level of 5%, we can reject $H_0$ when the 90% confidence interval around the observed estimate is in the predicted direction, and does not contain the value tested against (e.g., 0). In TOST, the first one-sided test is used to test against values smaller than the lower equivalence bound ($\Delta_{L}$), and the second one-sided test is used to test against values larger than the upper equivalence bound ($\Delta_{U}$). It is not necessary to control for multiple tests when using the TOST procedure because both tests need to be statistically significant to conclude equivalence. This means that when reporting a TOST result, authors can suffice by reporting the one-sided test with the smaller test parameter (e.g. $t$ or $r$) and thus the larger $p$-value. Statistical equivalence can be concluded when the largest of the two $p$-values is smaller than alpha.

Note that in this paper we will control the Type 1 error rate at 5% for all examples, mirroring the studies we reanalyze (but we recommend to justify the Type 1 error rate in original research based on substantive arguments [@Lakens2017b]. It might be easiest to think of an equivalence test as checking whether the confidence interval excludes the SESOI, but for any given study an equivalence test could also be conceptualized as determining whether effect sizes or test statistics are closer to zero than some critical value, or even whether the *p*-value for a NHST is larger than some *p*-value bound.

@@

In this article we provide five reproducible examples of equivalence tests which illustrate the procedure in easy-to-use free software (R, jamovi, and a spreadsheet). But first, we discuss different approaches to determining the SESOI for psychological research, which determines the statistical question an equivalence test answers.

@@

TOST is performed against equivalence bounds that are considered the smallest effect size of interest. The SESOI can sometimes be determined objectively, for example based on just noticeable differences. In lieu of objective justifications, the SESOI should ideally be based on a cost-benefit analysis. Since both costs and benefits are necessarily subjective, the SESOI will vary across researchers, fields, and time. The goal of setting a SESOI is to clearly justify why designing a study that has a high probability of rejecting effects more extreme than the specified equivalence bounds contributes to our knowledge base. A SESOI should be chosen such that inferences based on it answer a meaningful question. Although we use bounds that are symmetric around 0 in all examples in this manuscript for equivalence tests (e.g., $\Delta_{L} = -0.3$, $\Delta_{U} = 0.3$), it is also possible to use asymmetric bounds (e.g., $\Delta_{L} = -0.2$, $\Delta_{U} = 0.3$).

@@

Second, the SESOI can be based on published studies in the literature. Ideally, researchers who publish novel research would always specify their smallest effect size of interest, but this is not yet common practice. It is thus up to researchers who build on earlier work to decide which effect size is too small to be meaningful when examining the same hypothesis. @Simonsohn2015 recently proposed to set the SESOI to the effect size that an earlier study would have had 33% power to detect. With this so-called 'small telescopes' approach, the equivalence bounds are thus determined primarily based on the sample size in the original study. For example, consider a study in which `r N` participants answered a question, and the results were analyzed with a one-sample t-test. For a two-sided test with an alpha of `r alpha`, this test had 33% power to detect an effect of $d = `r printnum(pwr.t.test(n = N, sig.level = alpha, power = 0.33,type = "one.sample", alternative = "two.sided")$d)`$. Another example of how previous research can be used to determine the SESOI can be found in @Kordsmeyer2017, who defined the SESOI based on the mean of effect sizes reported in the literature. Thus, they examined whether their replication study could reject effects as large or larger than had on average been reported in the literature. Given random variation and bias in the literature, a more conservative approach could be to use the lower end of a confidence interval around the meta-analytic effect size estimate [cf. @Perugini2014].

Another justifiable choice when choosing the SESOI based on earlier work is to use the smallest observed effect size that could have been statistically significant in a previous study. In other words, we decide that effects that could not have yielded a $p < \alpha$ in an original study will not be considered meaningful in the replication study either, even if they were found to be statistically significant. Based only on the alpha level and the sample size, we can calculate the critical test value (e.g., *t*, *F*, *Z*). This critical test value can be transformed to a standardized effect size (e.g., $d = t \sqrt { \frac { 1} { n _ { 1} } + \frac { 1} { n _ { 2} } }$), which can thus be interpreted as the *critical effect size*.^[This will typically, although not always, correspond to the effect size the study had 50% power to detect [see @Lenth2007]. This procedure will result in equivalence bounds that are wider than the ones obtained using the small telescopes approach, which gives the effect size a study had 33% power to detect.] All observed effect sizes smaller than the critical effect size would not have been statistically significant in the original study, given the alpha and sample size of that study. By setting the SESOI to the critical effect size, an equivalence test can reject all observed effect sizes that could have been detected in an earlier study.

@@

Importantly, an equivalence test based on this approach does not answer any theoretical question (after all, the equivalence bounds are not based on any theoretical predictions) but it answers a *resource* question: Can effects be rejected that could reliably be detected given a specific sample size? In this case, concluding equivalence would suggest that effects which could reasonably be detected with a sample size of 100 can be rejected, and that future studies probably need larger sample sizes to examine a specific question. Whether or not the answer such an equivalence test provides is interesting depends on the sample sizes used in the study, and in part on how typical the chosen sample size is for the research area. An equivalence test based on 15 observations rejecting a SESOI of $d = 1$ cannot be expected to substantially increase our knowledge, given that most effects in psychology are substantially smaller than these bounds and require larger sample sizes to be detected. By transparently reporting the effects that can be detected and rejected based on the study design, researchers can communicate the information their study contributes, and provide a starting point for a discussion about what a reasonable SESOI may be.

@@

The SESOI, and thus the equivalence bounds, can be set in terms of standardized effect sizes (e.g. a Cohen’s $d$ of 0.5) or as a raw mean difference (e.g. 0.5 points on a 7-point scale). The key difference is that equivalence bounds set in raw differences are independent of the standard deviation, while equivalence bounds set as standardized effects are dependent on the standard deviation (since they are calculated as $\frac {X _ { 1} - X _ { 2} } {S D}$). The observed standard deviation randomly varies across samples. In practice, this means that when using standardized differences as bounds (e.g., $d = 0.35$), the equivalence test depends on the standard deviation in the sample. The equivalence bound for a raw mean difference of 0.5 equals a standardized equivalence bound of $d = 0.5$ when the standard deviation is 1, but a standardized equivalence bound of $d = 1$ when the standard deviation is 0.5.

@@

@Banerjee2012 reported that `r orig.N` participants who had been asked to describe an unethical deed from their past judged the room to be darker than participants who had been asked to describe an ethical deed ($t(`r orig.df`)= `r orig.t`$, $p= `r printp(orig.p)`$, $d= `r printnum(orig.d)`$). A close replication by @Brandt2014 with `r rep.n1+rep.n2` participants found no significant effect ($M_{unethical}= `r rep.m1`$, ${SD}_{unethical}= `r rep.sd1`$, $M_{ethical}=`r rep.m2`$, ${SD}_{ethical}=`r rep.sd2`$, $t(`r printnum(rep.df)`)=`r printnum(rep.t)`$, $p=`r printp(rep.p)`$, $d=`r printnum(rep.d)`$). Using the small telescopes approach, we can calculate the effect size the original study had 33% power to detect ($d = `r printnum(d.33)`$) and use this as the SESOI. A TOST procedure for Welch's *t*-test for independent samples and equivalence bounds of ${\Delta}_L=`r printnum(-d.33)`$ and ${\Delta}_U=`r printnum(d.33)`$ reveals that the effect observed in the replication study is statistically equivalent, $t(`r printnum(Brandt$TOST_df)`)=`r printnum(min(Brandt$TOST_t1, Brandt$TOST_t2))`$, $p=`r printp(max(Brandt$TOST_p1, Brandt$TOST_p2))`$ (Figure 2B).

@@

@Hyde2008 report that effect sizes for gender differences in mathematics tests across the 7 million students in the US represent trivial differences, which the authors specify as absolute effect sizes smaller than $d = `r printnum(sesoi)`$. For example, in grade 3 the difference is $d = 0.04$, with a standard error of 0.002. When we perform equivalence tests on the meta-analytic effect sizes of IQ difference for grades 2 to 11 (using an alpha level of `r printp(alpha)` to correct for multiple comparisons) and using equivalence bounds of $\Delta_{L} = -0.1$ and $\Delta_{U} = 0.1$, we see that all effect size estimates are measured with such high precision that they are statistically equivalent, and can be considered trivially small according to the authors’ definition (Figure 2C). In the case of grade 3, $z = `r printnum(max(gr3$TOST_Z1, gr3$TOST_Z2))`$, $p = `r printnum(max(gr3$TOST_p1, gr3$TOST_p2))`$. However, note that all of the effects are also statistically different from zero, as one might expect when there is no random assignment to conditions and samples sizes are very large. (e.g., for grade 3, $z = `r printnum(gr3.z)`$, *p* < 0.001. We see how equivalence tests allow researchers to distinguish between statistical significance and practical significance, which demonstrates how equivalence tests improve hypothesis testing procedures in psychological research.

@@

Since this is a replication, we could decide that we are interested in whether the observed effect size is smaller than the smallest effect size that the original study could have detected. This means that we use the critical z value (~1.96 in a two-tailed test with an alpha of 0.05) as equivalence bounds ($\Delta = \pm `r printnum(qnorm(1-0.05/2))`$). To calculate which difference corresponds to a critical z value in the original study, we multiply critical z value with the standard error ($`r printnum(crit.z)`$ * $`r printnum(se)`$), and find the original study could observe a significant effect for differences of $`r printnum(crit.diffe)`$ or more.

Since the original study had a clear directional hypothesis,  in this replication study we are only interested in whether receiving a warm pack *increases* the proportion of people who choose a gift for a friend.  This means we can test for inferiority (see Figure 1D) and conclude the absence of a meaningful effect if the observed effect size is reliably *lower* than the SESOI.
We find that we can reject effects larger than  $\Delta_{U} = `r printnum(crit.diff)`$, $z = `r printnum(Lynott$TOST_z2)`$, $p `r printp(Lynott$TOST_p2)`$ (see Figure 2D). Thus, we can conclude that the statistically non-significant effect is also statistically smaller than our SESOI.

@@

Kahane and colleagues [-@Kahane2015] investigate how responses to moral dilemmas in which participants have to decide whether or not they would sacrifice one person's life to save several other lives relate to other indicators of moral orientation. Traditionally, greater endorsement for sacrificing one life to save several others has been interpreted as a more "utilitarian" moral orientation (i.e. a stronger concern for the greater good). Kahane et al. contest this interpretation in a number of studies. In Study 4, they compare the traditionally used dilemmas with a set of new delimmas in which the sacrifice for the greater good is not another person’s life, but something the participant would have a partial interest in (e.g. buying a new mobile phone vs. donating the money to save lives in a distant country). The authors find no significant correlation between the two set of dilemmas, $r(`r Kahane.N-2`) = `r printnum(Kahane.r.GG)`$, $p = `r printp(Kahane.p)`$, $N = `r Kahane.N`$^[Study 4 in @Kahane2015 had a final sample size of $N = `r nrow(Kahane.data)`$, but due to missing data in `r sum(is.na(Kahane.data))` case, the correlation reported here is based on a sample of $N = `r Kahane.N`$.]. The authors conclude that “a robust association between ‘utilitarian’ judgment and genuine concern for the greater good seems extremely unlikely” (p. 206), given the statistical power their study had to detect meaningful effect sizes.

This interpretation can be formalized by performing an equivalence test for correlations, where equivalence bounds are set to an effect size the study had reasonable power to detect. With `r Kahane.N` participants, the study had 80% power to detect effects of $r = `r printnum(Kahane.r.80)`$. Performing the TOST procedure given bounds of $\Delta_{L} = `r printnum(-Kahane.r.80)`$ and $\Delta_{U} = `r printnum(Kahane.r.80)`$, the data is statistically equivalent, $r(`r Kahane.N-2`) = `r printnum(Kahane.r.GG)`$, $p = `r printp(max(Kahane$TOST_p1, Kahane$TOST_p2))`$. This means that values smaller than $r = `r printnum(-Kahane.r.80)`$ and larger than $r = `r printnum(Kahane.r.80)`$ can be rejected at an alpha level of 5%. If other researchers are interested in examining the presence of smaller effect size, they can design studies with a larger sample size.

@@

Equivalence testing is a simple statistical technique to reject the presence of a smallest effect size of interest. As long as we can calculate a confidence interval around a parameter estimate, we can compare it to the SESOI. The result of an equivalence test can be obtained by mere visual inspection of the CI [@Seaman1998; @Tryon2001], or by performing two one-sided tests (the TOST procedure).

As with any statistical test, the usefulness of the result of an equivalence test depends on the question we ask. The question manifests itself in the bounds we set: Is our data surprisingly close to zero, assuming a true effect exists that is as large as our smallest effect size of interest? If we test against very wide bounds, observing statistical equivalence can hardly be considered surprising, given that most effects in psychology are small to medium [@Hemphill2003]. 
Examining  papers citing @Lakens2017a, we found that some researchers state, but do not justify, the SESOI used in the equivalence test [e.g., @Brown2017; @Schumann2017]. An equivalence test using a SESOI of $d = 0.5$ might very well answer a question the researchers are interested in [for one possible justification based on minimally important differences, see @Norman2003], but researchers should always explain why they chose a SESOI. It makes little sense to report a statistical test, without explaining why one would want to answer this question to begin with. Equivalence bounds should be specified before results are known, ideally as part of a pre-registration [cf. @Piaggio2006]. Researchers might be tempted to set equivalence bounds *after* looking at their data, but like any other results-dependent decision, this "boundary hacking" will increase error rates.

We recommend that researchers by default perform both a null hypothesis significance test and an equivalence test on their data. The biggest challenge for researchers will be to specify the smallest effect size of interest. Psychological theories are usually too vague to derive precise predictions, and if there are no theoretical reference points, natural constraints, or prior studies a researcher can use to define the SESOI for a hypothesis test, any choice will be somewhat arbitrary. In some research lines researchers might simply use equivalence tests to reject consecutively smaller effect sizes by performing studies with increasingly larger sample sizes while controlling error rates, until no one is willing to invest the time and resources needed to examine the presence of even smaller effects. Nevertheless, it is important to realize that not specifying a SESOI for our research questions at all will severely hinder theoretical progress [@Morey2016]. Incorporating equivalence tests in our statistical toolbox will in time contribute to better --- and more falsifiable --- theories.
